SUPPLEMENTARY:\\Correcting for bias in high-throughput sequencing data

### 2011-07-08 seqbias-paper.tex
small typo
move figures, to placate bioinformatics upload tool
final revisions before submission

\title{Correcting for bias in high-throughput sequencing data}
\author[Jones \textit{et~al}]
{Daniel C. Jones\,$^{1,}$
\footnote{to whom correspondence should be addressed}\hspace{0.5em}
Walter L. Ruzzo\,$^{1,2}$
Xinxia Peng\,$^{3}$
Michael G. Katze$^{3}$
}

@@

\address{
$^{1}$Department of Computer Science and Engineering, University of
Washington, Seattle, WA 98195-2350, USA\\
$^{2}$Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA\\
$^{3}$Department of Microbiology, University of Washington, Seattle, WA
98195-7242, USA}

@@

These biases may adversely effect transcript discovery, as low level noise may
be over reported in some regions, and in others, active transcription may be
under reported. They render untrustworthy comparisons of relative abundance
between genes or isoforms, and any test of differential expression hangs on the
assumption that these biases are identical between replicates.  Additionally, in
many tests of differential expression---similar to the effects of transcript
length---a higher read count will result in higher statistical confidence. It
follows that the sensitivity of such a test will also be biased by sequence,
affecting downstream analysis such as gene ontology enrichment tests.

@@

\begin{figure*}
\centerline{\includegraphics[width=0.8\textwidth]{fig/freqs.eps}}
\caption{Nucleotide frequencies are plotted relative to the start (labeled,
position 0) of each mapped read, respecting strand. The sequence is taken from
the genomic context surrounding the read, so that -40 to -1, for example, fall
outside the read sequence itself. The symmetrized Kullback-Leibler divergence is
used to summarize the difference in nucleotide frequency compared to a fixed
estimate of background nucleotide frequencies made by sampling many positions
nearby mapped reads.  Under the assumption that reads are sampled uniformly from
transcripts, each of the plots should be essentially flat.}
\label{fig:freqs}
\end{figure*}

\citet{Li2010} propose two models. The first is a Poisson linear
model, in which read counts across a transcript follow inhomogeneous Poisson
process. The read count at position $i$ within the transcript is Poisson
distributed with parameter $\lambda_i$, where, $\log(\lambda_i)$ is the sum of
weights determined by the nucleotide at each position surrounding the read
start, in addition to a term capturing the abundance of the transcript.

@@

Another model, proposed by \citet{Hansen2010}, directly estimates the
distribution of initial heptamers within reads, then estimates a presumed
background heptamer distribution, sampled from the end of reads. The read count
at a given position is then adjusted by the ratio of the foreground and
background heptamer probabilities. Specifying two distributions over heptamers
(i.e. foreground and background distributions) requires $2(4^7-1) = 32,766$
parameters, so while no gene annotations are needed to train such a model, a
significant number of reads are required.

Lastly, \citet{Roberts2011} have very recently published a description of
another approach, in which sequence probabilities are modeled by variable-order
Markov chains. The structure of these Markov chains are hard-coded, chosen in
advance using a hill-climbing algorithm on a representative data set. This method
is implemented in the latest version of Cufflinks \citep{Trapnell2010}, and
tightly incorporated into its estimation of transcript abundance.

Here we propose a new approach, using Bayesian networks to model sequence
probabilities. Unlike the methods of Roberts or Li, our model requires no gene
annotations, nor even the assumption that the short reads are derived from RNA.
In this sense, we build on the work done by \citet{Hansen2010}, generalizing
their approach in a way we find to be more robust and effective at correcting
for bias in a variety of protocols. Due to the weak assumptions required my our
model, it is applicable and potentially useful in any setting in which short
reads are aligned to a reference sequence.

@@

Towards that end, we propose training a Bayesian network classifier trained on
examples of foreground and background sequences. Though our goal is not
classification, by using the machinery of classification we can avoid a model
that is over-parametrized. Parameters that are not informative in
discriminating between foreground and background are not included in the model.
The Bayesian network can then used to evaluate sequence probability, and thus
bias, at any genomic position. Figure \ref{fig:overview} gives a brief overview
of the proposed model.

\begin{figure}
\centerline{\includegraphics[width=0.35\textwidth]{fig/overview.eps}}
\caption{An overview of the approach taken: (a) foreground sequences are sampled
from the regions surrounding the starts of mapped reads, (b) background
sequences are sampled by randomly offsetting foreground positions, (c) a
Bayesian network is classifier is trained to discriminate between the set of
sampled foreground and background sequences, (d) and the classifier is evaluated
at each position within a locus, predicting bias.}
\label{fig:overview}
\end{figure}

However, we have so for ignored one complication: the RNA abundance that we wish
to estimate is not itself independent of the nucleotide sequence. Notably,
exonic DNA tends to be more GC-rich than intergenic DNA. If background sequences
are sampled uniformly from the genome we run the risk of incorrectly adjusting
for biological sequence bias, rather than technical sequence bias.  To avoid
this, we propose using paired training data. Each foreground training
sequence is paired with a background sequence taken from nearby position
that is likely to have similar abundance and general nucleotide composition.
Alternatively, we could pair foreground samples with background samples from
within the same transcript, but we prefer to avoid dependence on existing
gene annotations as such a model would be limited to RNA-Seq analysis in which
trustworthy and reasonably complete gene annotations are available.

The methods proposed \citet{Hansen2010} and \cite{Roberts2011} also treat bias
correction as a problem of estimating foreground and background sequence
probabilities. They differ primarily in how these sequence probabilities are
estimated. \citet{Li2010} estimate reweighting coefficients ($b_i$, in our
notation) directly, given training data consisting of long, highly expressed
transcripts.

@@

In our model, we do not rely on constraining the set of networks (e.g. to
trees), and instead approximate the NP-Hard problem of determining the optimal
network structure using a fast hill-climbing algorithm. Furthermore, we
train our model \emph{discriminatively}; only parameters that are deemed
informative in discriminating between foreground and background sequences are
included in the model. We thus seek to train a model that reduces bias, without
including uninformative parameters that would only increase variance.

@@

The model is trained on $n$ sequences, one half labeled as foreground, the other
background, sampled from the reference genome. To obtain the foreground
sequences, we take sequences surrounding the starts positions of $n/2$ aligned
reads. We wish to avoid overfitting to any short locus with a large abundance of reads,
yet we must also capture the bias introduced by PCR amplification. We therefore
choose the $n/2$ unique reads with the highest number of duplicates,
yet count each read only once, ignoring duplicates

To obtain background training sequences, we randomly offset the positions from
which the foreground sequences were sampled.  The offset is drawn from a
zero-mean Gaussian, and rounded to the nearest integer, away from zero.
By using such a scheme, we attempt to mitigate the effects of biological
sequence bias, sampling positions that are more likely to be biologically
similar.

@@

To determine the structure and parameters of the Bayesian network, we use a
hill-climbing approach similar to the algorithm described by \citet{Grossman2004}.
The network structure is determined by greedily optimizing the conditional
log-likelihood.
\begin{align*}
\ell &= \sum_{i=1}^{n} \log \Pr( x_i | s_i ) \\
&=
\sum_{i=1}^{n} \log \frac{ \Pr(s_i | x_i) \Pr( x_i ) }{
\sum_{x \in \{0,1\}} \Pr( s_i | x ) \Pr(x) } \\
\end{align*}

Where $\Pr(x)$ is flat (i.e.  $\Pr( x = 0 ) = \Pr( x = 1 ) =
0.5$) if we sample foreground and background positions equally.

@@

\begin{figure*}
\centerline{\includegraphics[width=\textwidth]{fig/models.eps}}
\caption{
The network structures learned on each of the data sets are examined. Positions
are relative to the read start, which is labeled 0. Hollow circles indicate
positions that were not included in the model, being deemed uninformative, given
the other positions and edges. The number of parameters needed to
specify each model is listed in parenthesis below.}
\label{fig:models}
\end{figure*}

@@

\begin{table}
\processtable{Data sets on which the methods are evaluated.\label{tab:datasets}}
{
\begin{tabular}{lllr}\toprule
Experiment & Species & Platform & Read Length \\\midrule
\citet{Wetterbom2010} & Chimpanzee & AB & 33 \\
Katze (unpublished) & Macaque & AB & 50 \\
\citet{Bullard2010} & Human & Illumina & 35 \\
\citet{Mortazavi2008} & Mouse & Illumina & 33 \\
\citet{Trapnell2010} & Mouse & Illumina & 75 \\\botrule
\end{tabular}
}{}
\end{table}

@@

Our own method was trained on the 100,000 reads from chromosomes 1--8, with the
highest number of duplicates.

All data sets were mapped using Bowtie \citep{Langmead2009} against the latest
genome assemblies available from the UCSC Genome Browser \citep{Karolchik2008}
during preparation.

@@

We compute the divergence by reweighting the read counts using the predicted
bias coefficient, choosing the upper half in each exon, discarding duplicate
reads, and then tallying frequencies of overlapping $k$-mers. The $k$-mer
distribution obtained are then compared to a background distribution obtained by
redistributing reads uniformly at random within their exons.

We repeated the procedure for $k \in \{1, 2, 3, 4, 5, 6\}$. The results of this
analysis are plotted in Figure \ref{fig:kl}, for $k = 1$ and $k = 4$. The
remaining cases are plotted in Supplementary Section 4.

\begin{figure*}
\centerline{\includegraphics[width=1.00\textwidth]{fig/kl.eps}}
\caption{The Kullback-Leibler divergence compares the frequency of $k$-mers
(here, for $k = 1$ and $k = 4$) surrounding the starts of aligned reads to the
frequencies expected under the assumption of uniform sampling from within exons.
A large divergence indicates significant bias. Plotted here is the divergence
from unadjusted reads counts as well as after adjusting read counts using each
method.}
\label{fig:kl}
\end{figure*}

@@

This measure can be interpreted as the improvement in fit over the null model,
with $R^2 = 1$ indicating a prefect fit, and a negative numbers, an increasingly
worse fit. A more dubious, yet intuitive interpretation of $R^2$ is the
proportion of variability enplaned by the model.

@@

\begin{figure}
\centerline{\includegraphics[width=0.3\textwidth]{fig/pois-boxplot.eps}}
\caption{For each of the 1000 test exons, we measure a pseudo-coefficients of
determination $R^2$, equivalent to the improvement in log-likelihood under the
bias corrected model. The statistic is positive when uniformity is increased,
and negative when decreased. Boxes are plotted to mark the 25\%, 50\%, and 75\%
quantiles, with whiskers extending to 1.5 times the inter-quartile range (i.e.,
the span between the 25\% and 75\% quantiles).
}

@@

\begin{table}
\processtable{
    The median $R^2$ goodness fo fit statistic across test exons.
\label{tab:pois}}{
    \begin{tabular}{lcccc}\toprule
          & BN    & MART  & GLM   & 7mer \\\midrule
Wetterbom & 0.174 & 0.016 & 0.066 & -0.079 \\
Katze     & 0.280 & 0.243 & 0.158 & 0.033 \\
Bullard   & 0.267 & 0.163 & 0.224 & 0.157 \\
Mortazavi & 0.240 & 0.210 & 0.197 & 0.091 \\
Trapnell  & 0.289 & 0.289 & 0.248 & 0.138
    \end{tabular}
}
{
    The $R^2$ statistic measures increased uniformity in read coverage, after
    correcting for bias.  Here the median $R^2$ across the test exons is listed
    for each method and sample. A positive $R^2$ indicates a better fit.
}
\end{table}

@@

In our results, estimating initial heptamer frequencies was not seen to be as
effective as the other models, even when data generated using random hexamer
priming was used. Given the large number of parameters needed to estimate
heptamer frequencies, a likely explanation is that the model is overfit to the
training set. Though the training set consisted of at least 600,000 reads in
each experiment, it seems more reads, or a greater diversity of reads are needed
to produce an accurate or useful estimate.

@@

We have withheld any direct comparison to the method described by
\citet{Roberts2011} and implemented in Cufflinks \citep{Trapnell2010}. Though
this method is superficially similar to our own, a proper comparison is
difficult, as the software cannot be applied independently of estimating
transcript abundance in FPKM (fragments per exonic kilobase, per million mapped
reads) using Cufflinks. Fairness would dictate that competing methods be
substituted in FPKM estimation, or that a separate interface be written to the
Cufflinks bias correction method---both comparisons requiring significant effort.

Though the Cufflinks method and our own both use graphical models to estimate
sequence probabilities, we make no restriction on the graph other than
acyclicity. We go to considerable effort to efficiently approximate the optimal
structure for each data set rather than using a fixed structure, as in
Cufflinks. A ``one size fits all'' approach likely works quite well in many
cases, yet the observed specificity of the bias to protocol and platform argues
against it. For example, the structures learned by our method (Figure
\ref{fig:models}) are considerably different between those sequenced on an
Illumina platform versus an AB platform.

Training our model on more reads leads to more accurate estimation of bias, but
an increasingly long training time. For example, in our tests, fitting our model to
100,000 reads from the Mortazavi data, training time was approximately 45
minutes, running on one core of a 3 Ghz Intel Xeon processor. However, limiting
the training to 25,000 reads leads to a model that is nearly as accurate while
requiring under 4 minutes to train. A full discussion of this trade-off is in
Supplementary Section 5.

Because we do not require annotations, ChIP-Seq, and other high-throughput
sequencing experiments, may also benefit from our model. In a preliminary
investigation, we found the sequence bias in one ChIP-Seq experiment
\cite{Cao2010} was less than that observed in any of the RNA-Seq data we
evaluated; however, our method is still able to effectively correct for the bias
that was observed (see, Supplementary Section 2).
Protocol differences, as we have seen,
can result in significant differences in observed nucleotide frequencies, so we
may not safely assert that bias in ChIP-Seq data is always low.  Given the weak
assumptions made by our model, our estimation of bias could easily be
incorporated into ChIP-Seq peak-calling algorithms, and potentially improve
accuracy.

RNA-Seq is most often used to compare levels of expression, and so a natural
concern is the consistency of the bias between samples. In the data we examined,
the bias appears to be largely, but not entirely consistent (see, Supplementary
Section 3).  Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced
on the Illumina platform display similar patterns of nonuniformity, yet differ
in magnitude, suggesting that batch effects in RNA-Seq remain a legitimate
concern that should not be dismissed without evaluation.

>>>>

\title{Correcting for bias in high-throughput sequencing data}
\author[Jones \textit{et~al}]
{Daniel C. Jones\,$^{1,}$
\footnote{to whom correspondence should be addressed}\hspace{0.5em}
Walter L. Ruzzo\,$^{1,2,3}$
Xinxia Peng\,$^{4}$
Michael G. Katze$^{4}$
}

@@

\address{
$^{1}$Department of Computer Science and Engineering, University of
Washington, Seattle, WA 98195-2350, USA\\
$^{2}$Department of Genome Sciences, University of Washington, Seattle, Wa
98195-5065, USA\\
$^{3}$Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA\\
$^{4}$Department of Microbiology, University of Washington, Seattle, WA
98195-7242, USA}

@@

These biases may adversely effect transcript discovery, as low level noise may
be over-reported in some regions, and in others, active transcription may be
under-reported. They render untrustworthy comparisons of relative abundance
between genes or isoforms, and any test of differential expression hangs on the
assumption that these biases are identical between replicates.  Additionally, in
many tests of differential expression---similar to the effects of transcript
length---a higher read count will result in higher statistical confidence. It
follows that the sensitivity of such a test will also be biased by sequence,
affecting downstream analysis such as gene ontology enrichment tests.

@@

\begin{figure*}
\centerline{\includegraphics[width=0.9\textwidth]{freqs.eps}}
\caption{Nucleotide frequencies are plotted relative to the start (labeled
position 0) of each mapped read, respecting strand, and grouped by platform
(Illumina or ABI SOLiD).  The data sets plotted here are those used
for evaluation, listed in Table \ref{tab:datasets}.
The sequence is taken from the genomic context surrounding the read, so that -40
to -1, for example, fall outside the read sequence itself. The symmetrized
Kullback-Leibler divergence is used to summarize the difference in nucleotide
frequency compared to a fixed estimate of background nucleotide frequencies made
by sampling many positions near mapped reads.  Under the assumption that reads
are sampled uniformly from transcripts, each of the plots should be essentially
flat}
\label{fig:freqs}
\end{figure*}

\citet{Li2010} propose two models. The first is a Poisson linear
model, in which read counts across a transcript follow an inhomogeneous Poisson
process. The read count at position $i$ within the transcript is Poisson
distributed with parameter $\lambda_i$, where, $\log(\lambda_i)$ is the sum of
weights determined by the nucleotide at each position surrounding the read
start, in addition to a term capturing the abundance of the transcript.

@@

Another model, proposed by \citet{Hansen2010}, directly estimates the
distribution of initial heptamers within reads, then estimates a presumed
background heptamer distribution, sampled from the ends of reads. The read count
at a given position is then adjusted by the ratio of the foreground and
background heptamer probabilities. Specifying two distributions over heptamers
(i.e. foreground and background distributions) requires \mbox{$2(4^7-1) =
32,766$}
parameters, so while no gene annotations are needed to train such a model, a
significant number of reads are required.

Lastly, \citet{Roberts2011} have very recently published a description of
another approach, in which sequence probabilities are modeled by variable-order
Markov chains. The structure of these Markov chains are hard-coded, chosen in
advance using a hill-climbing algorithm on a representative data set. This method
is implemented in the latest version of Cufflinks \citep{Trapnell2010}, and
tightly incorporated into its estimation of transcript abundance, requiring
either predicted or existing gene annotations.

Here we propose a new approach, using Bayesian networks to model sequence
probabilities. Unlike the methods of Roberts or Li, our model requires no gene
annotations, nor even the assumption that the short reads are derived from RNA.
In this sense, we build on the work done by \citet{Hansen2010}, generalizing
their approach in a way we find to be more robust and effective at correcting
for bias in a variety of protocols. Due to the weak assumptions required by our
model, it is applicable and potentially useful in any setting in which short
reads are aligned to a reference sequence.

@@

Towards that end, we propose training a Bayesian network classifier trained on
examples of foreground and background sequences. Though our goal is not
classification, by using the machinery of classification we can avoid a model
that is over-parametrized. Parameters that are not informative in
discriminating between foreground and background are not included in the model.
The Bayesian network can then be used to evaluate sequence probability, and thus
bias, at any genomic position. Figure \ref{fig:overview} gives a brief overview
of the proposed model.

\begin{figure}
\centerline{\includegraphics[width=0.35\textwidth]{overview.eps}}
\caption{An overview of the approach taken: (a) foreground sequences are sampled
from the regions surrounding the starts of mapped reads, (b) background
sequences are sampled by randomly offsetting foreground positions, (c) a
Bayesian network is classifier is trained to discriminate between the set of
sampled foreground and background sequences, (d) and the classifier is evaluated
at each position within a locus, predicting bias.}
\label{fig:overview}
\end{figure}

However, we have so for ignored one complication: the RNA abundance that we wish
to estimate is not itself independent of the nucleotide sequence. Notably,
exonic DNA tends to be more GC-rich than intergenic DNA. If background sequences
are sampled uniformly from the genome we run the risk of incorrectly adjusting
for biological sequence bias, rather than technical sequence bias.  To avoid
this, we propose using paired training data. Each foreground training
sequence is paired with a background sequence taken from a nearby position
that is likely to have similar abundance and general nucleotide composition.
Alternatively, we could pair foreground samples with background samples from
within the same transcript, but we prefer to avoid dependence on existing
gene annotations as such a model would be limited to RNA-Seq analysis in which
trustworthy and reasonably complete gene annotations are available.

The methods proposed by \citet{Hansen2010} and \cite{Roberts2011} also treat bias
correction as a problem of estimating foreground and background sequence
probabilities. They differ primarily in how these sequence probabilities are
estimated. \citet{Li2010} estimate reweighting coefficients ($b_i$, in our
notation) directly, given training data consisting of long, highly expressed
transcripts.

@@

In our model, we do not rely on constraining the set of networks (e.g., to
trees), and instead approximate the NP-Hard problem of determining the optimal
network structure using a fast hill-climbing algorithm. Furthermore, we
train our model \emph{discriminatively}; only parameters that are deemed
informative in discriminating between foreground and background sequences are
included in the model. We thus seek to train a model that reduces bias, without
including uninformative parameters that would only increase variance.

@@

The model is trained on $n$ sequences, one half labeled as foreground, the other
background, sampled from the reference genome. To obtain the foreground
sequences, we take sequences surrounding (extending 20 nt to either side, by
default) the starts positions of $n/2$ aligned reads. We wish to avoid
overfitting to any short locus with a large abundance of reads, yet we must also
capture the bias introduced by, for exapmle, PCR amplification. We therefore
choose the $n/2$ unique reads with the highest number of duplicates, yet count
each read only once, ignoring duplicates

To obtain background training sequences, we randomly offset the positions from
which the foreground sequences were sampled.  The offset is drawn from a
zero-mean Gaussian (with $\sigma^2 = 10$, by default), and rounded to the
nearest integer, away from zero.  By using such a scheme, we attempt to mitigate
the effects of biological sequence bias, sampling positions that are more likely
to be biologically similar.

@@

To determine the structure and parameters of the Bayesian network, we use a
hill-climbing approach similar to the algorithm described by \citet{Grossman2004}.
The network structure is determined by greedily optimizing the conditional
log-likelihood:
\begin{align*}
\ell &= \sum_{i=1}^{n} \log \Pr( x_i | s_i ) \\
&=
\sum_{i=1}^{n} \log \frac{ \Pr(s_i | x_i) \Pr( x_i ) }{
\sum_{x \in \{0,1\}} \Pr( s_i | x ) \Pr(x) } \\
\end{align*}
where $\Pr(x)$ is flat (i.e.  $\Pr( x = 0 ) = \Pr( x = 1 ) =
0.5$) if we sample foreground and background positions equally.

@@

\begin{figure*}
\centerline{\includegraphics[width=\textwidth]{models.eps}}
\caption{
The network structures learned on each of the data sets are displayed. Positions
are relative to the read start, which is labeled 0. Hollow circles indicate
positions that were not included in the model, being deemed uninformative, given
the other positions and edges. The number of parameters needed to
specify each model is listed in parenthesis below.}
\label{fig:models}
\end{figure*}

@@

\begin{table}
\processtable{Data sets on which the methods are evaluated.\label{tab:datasets}}
{
\begin{tabular}{lllr}\toprule
Experiment & Species & Platform & Read Length \\\midrule
\citet{Wetterbom2010} & Chimpanzee & ABI & 33 \\
Katze (unpublished) & Macaque & ABI & 50 \\
\citet{Bullard2010} & Human & Illumina & 35 \\
\citet{Mortazavi2008} & Mouse & Illumina & 33 \\
\citet{Trapnell2010} & Mouse & Illumina & 75 \\\botrule
\end{tabular}
}{}
\end{table}

@@

Our own method was trained on the 100,000 reads from chromosomes 1--8, with the
highest number of duplicate reads.

All data sets were mapped using Bowtie \citep{Langmead2009} against,
respectively, the hg19, mm9, and panTro2 genome assemblies obtained from the
UCSC Genome Browser \citep{Karolchik2008}.

@@

We compute the divergence by reweighting the read counts using the predicted
bias coefficient, choosing the upper half in each exon, discarding duplicate
reads, and then tallying frequencies of overlapping $k$-mers. The $k$-mer
distribution obtained is then compared to a background distribution obtained by
redistributing reads uniformly at random within their exons.

We repeated the procedure for $k \in \{1, 2, 3, 4, 5, 6\}$. The results of this
analysis are plotted in Figure \ref{fig:kl}, for $k = 1$ and $k = 4$. The
remaining cases are plotted in Supplementary Section 2.

\begin{figure*}
\centerline{\includegraphics[width=1.00\textwidth]{kl.eps}}
\caption{The Kullback-Leibler divergence compares the frequency of $k$-mers
(here, for $k = 1$ and $k = 4$) surrounding the starts of aligned reads to the
frequencies expected under the assumption of uniform sampling from within exons.
A large divergence indicates significant bias. Plotted here is the divergence
from unadjusted read counts as well as after adjusting read counts using each
method.}
\label{fig:kl}
\end{figure*}

@@

This measure can be interpreted as the improvement in fit over the null model,
with $R^2 = 1$ indicating a prefect fit, and smaller numbers, an increasingly
worse fit. A more dubious, yet intuitive interpretation of $R^2$ is the
proportion of variability explained by the model.

@@

\begin{figure}
\centerline{\includegraphics[width=0.35\textwidth]{pois-boxplot.eps}}
\caption{For each of the 1000 test exons, we compute a pseudo-coefficients of
determination $R^2$, equivalent to the improvement in log-likelihood under the
bias corrected model. The statistic is positive, and increases as uniformity is
increased, and negative when uniformity is decreased. Boxes are plotted to mark
the 25\%, 50\%, and 75\% quantiles, with whiskers extending to 1.5 times the
inter-quartile range (i.e., the span between the 25\% and 75\% quantiles), and
dots marking more extreme values.
}

@@

\begin{table}
\processtable{
    The median $R^2$ goodness of fit statistic across test exons.
\label{tab:pois}}{
    \begin{tabular}{lcccc}\toprule
          & BN    & MART  & GLM   & 7mer \\\midrule
Wetterbom & 0.174 & 0.016 & 0.066 & -0.079 \\
Katze     & 0.280 & 0.243 & 0.158 & 0.033 \\
Bullard   & 0.267 & 0.163 & 0.224 & 0.157 \\
Mortazavi & 0.240 & 0.210 & 0.197 & 0.091 \\
Trapnell  & 0.289 & 0.289 & 0.248 & 0.138
    \end{tabular}
}
{
    The $R^2$ statistic measures increased uniformity in read coverage, after
    correcting for bias.  Here the median $R^2$ across the test exons is listed
    for each method and sample. A higher $R^2$ indicates a better fit.
}
\end{table}

@@

In our results, estimating initial heptamer frequencies was not seen to be as
effective as the other models, even when data generated using random hexamer
priming was used. Given the large number of parameters needed to estimate
heptamer frequencies, a likely explanation is that the model is overfit to the
training set. Though we trained the 7mer model on at least 600,000 reads in each
experiment, it seems more reads, or a greater diversity of reads are needed
to produce an accurate or useful estimate.

@@

We have not performed any direct comparison to the method described by
\citet{Roberts2011} and implemented in Cufflinks \citep{Trapnell2010}. Though
this method is superficially similar to our own, a proper comparison is
difficult, as the software cannot be applied independently of estimating
transcript abundance in FPKM (fragments per exonic kilobase, per million mapped
reads) using Cufflinks. Fairness would dictate that competing methods be
substituted in FPKM estimation, or that a separate interface be written to the
Cufflinks bias correction method---both comparisons requiring significant effort.

Though the Cufflinks method and our own both use graphical models to estimate
sequence probabilities, we make no restriction on the graph other than
acyclicity. We go to considerable effort to efficiently approximate the optimal
structure for each data set rather than using a fixed structure, as in
Cufflinks. A ``one size fits all'' approach likely works quite well in many
cases, yet the observed specificity of the bias to protocol and platform argues
against it. For example, the structures learned by our method (Figure
\ref{fig:models}) are considerably different between those sequenced on an
Illumina platform versus an ABI platform.

Training our model on more reads leads to more accurate estimation of bias, but
an increasingly long training time. For example, in our tests, fitting our model to
100,000 reads from the Mortazavi data, training time was approximately 45
minutes, running on one core of a 3 Ghz Intel Xeon processor. However, limiting
the training to 25,000 reads leads to a model that is nearly as accurate while
requiring less than 4 minutes to train. A full discussion of this trade-off is in
Supplementary Section 3.

Because we do not require annotations, ChIP-Seq, and other high-throughput
sequencing experiments, may also benefit from our model. In a preliminary
investigation, we found the sequence bias in one ChIP-Seq experiment
\cite{Cao2010} was less than that observed in any of the RNA-Seq data we
evaluated; however, our method is still able to effectively correct for the bias
that was observed (see, Supplementary Section 4).
Protocol differences, as we have seen,
can result in significant differences in observed nucleotide frequencies, so we
may not safely assert that bias in ChIP-Seq data is always low.  Given the weak
assumptions made by our model, our estimation of bias could easily be
incorporated into ChIP-Seq peak-calling algorithms, and potentially improve
accuracy.

RNA-Seq is most often used to compare levels of expression, and so a natural
concern is the consistency of the bias between samples. In the data we examined,
the bias appears to be largely, but not entirely consistent (see, Supplementary
Section 5).  Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced
on the Illumina platform display similar patterns of nonuniformity, yet differ
in magnitude, suggesting that batch effects in RNA-Seq remain a legitimate
concern that should not be dismissed without evaluation.

### 2011-07-08 seqbias-supplement.tex
various corrections from markus korth
final revisions before submission

%\setmainfont{Helvetica}

@@

%\maketitle

\dcjtitle{\sc{(SUPPLEMENTARY)}}{\sc{A Model for Sequencing Bias in RNA-Seq}}{Daniel C. Jones}

%\begin{center}
%\large{
%\sc{SUPPLEMENTARY}

%\sc{A model for Sequence Bias in RNA-Seq}
%}
%\end{center}

@@

Observing the nonuniform distribution of nucleotide frequencies surrounding the
5' end reads, a natural step to take would be to trim the 5' end before mapping,
in the hope that simply removing the portion of the read in which the bias
occurs will also remove the bias. The figure below demonstrates that this is not
the case. Trimming the initial heptamer in the Mortazavi data set does nothing to
reduce the bias, and simply shifts the plot by seven positions. This indicates
that the issue is \emph{sampling bias}, rather than a bias in base calling.

@@

We did however repeat our analysis using the Kullback-Leibler divergence.
We used several variations of the method described by Hansen, et. al.,
\cite{Hansen2010}. ``7mer'' estimated the initial heptamer, ``Avg-7mer''
averages the initial two heptamers, and ``4mer'' averages the initial 4mer.

>>>>

@@

\dcjtitle{\sc{(SUPPLEMENTARY)}}{\sc{Correcting for Bias in High-Throughput
Sequencing Data}}{Daniel C. Jones}

@@

Observing the nonuniform distribution of nucleotide frequencies surrounding the
5' end of reads, a natural step to take would be to trim the 5' end before mapping,
in the hope that simply removing the portion of the read in which the bias
occurs will also remove the bias. The figure below demonstrates that this is not
the case. Trimming the initial heptamer in the Mortazavi data set does nothing to
reduce the bias, and simply shifts the plot by seven positions. This indicates
that the issue is \emph{sampling bias}, rather than a bias in base calling.

@@

We did however repeat our analysis using the Kullback-Leibler divergence.
We used several variations of the method described by Hansen, et. al.,
\cite{Hansen2010}. ``7mer'' estimated the initial heptamer, ``Avg-7mer''
averages the initial two heptamers, and ``4mer'' averages the initial two 4mers.
